Response to reviewer

### 2025-08-01 paper_plos/paper_plos_V0.tex
Made the existing draft as concise as possible -- shed 6 pages. Old version is paper_plos_V0.qmd and .pdf.

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\usepackage{colortbl}
\usepackage{array}
\usepackage{anyfontsize}
\usepackage{multirow}
\usepackage{float}
\floatplacement{table}{H}
\floatplacement{figure}{H}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={A family of accessibility measures derived from spatial interaction principles},
  pdfauthor={Anastasia Soukhov; Rafael H. M. Pereira; Christopher D. Higgins; Antonio Páez},
  pdfkeywords={accessibility; spatial interaction model; balancing
factors; competitive accessibility; constraints; singly constrained;
doubly constrained; total constrained},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{A family of accessibility measures derived from spatial
interaction principles}
\author{Anastasia Soukhov \and Rafael H. M. Pereira \and Christopher D.
Higgins \and Antonio Páez}
\date{}

\begin{document}
\maketitle
\begin{abstract}
Transportation planning has long prioritized the efficiency of movement,
or mobility. However, the concept of accessibility represents a more
comprehensive evolution, shifting focus from mere movement to the
potential to reach (i.e., spatially interact) with desired destinations.
Despite growing recognition of accessibility-based planning approaches,
the concept remains fragmented, with inconsistent definitions and
unclear interpretations. This work's aim is to clarify and unify the
concept of accessibility by connecting it into spatial interaction
modeling. We demonstrate that widely used mobility and accessibility
models, such as gravity-based accessibility and spatial interaction
models, share common theoretical roots. From this foundation, this paper
offers three contributions: (A) we introduce a family of accessibility
measures within the principles of spatial interaction, and (B) formally
define four members of the family, namely the `unconstrained' measure
(i.e., Hansen-type accessibility), the `total constrained' measure
(i.e., a constrained version of the Hansen-type accessibility), the
`singly constrained' measure (i.e., related to the popular two step
floating catchment approach - 2SFCA), and the `doubly constrained'
measure representing realized interactions or `access', effectively
equal to the doubly constrained spatial interaction model; and (C) we
demonstrate the interpretability advantages of the family, as these
constrained accessibility measures yield values in units of the number
of potential ``opportunities for spatial interaction'' or ``population
for spatial interaction'' for each zone and zonal flow. The family of
accessibility measures proposed here clarifies the concept of
`potential' in accessibility, demonstrates theoretical and formulaic
linkages across popular accessibility and spatial interaction models,
and reintroduces measurement units into accessibility measures. By doing
so, we believe this family of measures can unlock a clearer, more
interpretable, and cohesive foundation for accessibility analysis.
\end{abstract}

@@

In the early nineteenth century, industrializing cities grew rapidly: so
did traffic congestion and the development of transportation planning
practice focused primarily on mobility. In this newly founded practice,
access to destinations was treated as a by-product of movement.
Following the rise of the automobile and significant investments in
transportation infrastructure after World War II, this mobility-oriented
approach has led to some problematic outcomes. Specifically, the car
became seen as the ultimate mobility tool, helping to foster the
development of low-density, single-use residential neighborhoods and
entrenching an automobility mono-culture within transportation systems
(H. J. Miller 2011; Lavery, Páez, and Kanaroglou 2013). Decades of
planning for this automobility mono-culture, still often characterized
by road and highway expansion, have been marked by increased travel
costs and environmental burdens, with limited impact on enhancing the
ease with which people can reach destinations (Steven Farber and Páez
2011; S. Handy 2002; Páez et al. 2010). In response, transportation
researchers have increasingly advocated for the adoption of
accessibility as a planning criterion, in contrast to traditional
mobility-oriented transportation planning approaches which translate
into indicators that benchmark movement (e.g., vehicle kilometres
traveled, intersection through traffic, etc.) which are not necessarily
linked to improved accessibility (Silva et al. 2017; Paez et al. 2013;
S. Handy 2020; El-Geneidy and Levinson 2022).

Accessibility may be defined as the ``potential of opportunities for
{[}spatial{]} interaction'' (Hansen 1959), in contrast to mobility which
is simply, movement in itself or `spatial interaction'. Mobility has
been the basis of traditional transportation planning approaches
(Ortúzar and Willumsen 2011). While accessibility brings a more holistic
understanding of combined transportation and land use systems,
incorporating both the explicit consideration of destination as
opportunity and the concept of \emph{potential} (S. L. Handy and
Niemeier 1997).

The growing interest in accessibility has been accompanied by a boom in
scholarly research using different methods and focusing on different
research contexts; it has grown to include studies of access to
employment (e.g., Karst and Van Eck 2003; Grengs 2010; Páez et al. 2013;
Merlin and Hu 2017; Tao et al. 2020), health care (e.g., Luo and Wang
2003; Páez et al. 2010; Wan, Zou, and Sternberg 2012; Delamater 2013;
Boisjoly, Moreno-Monroy, and El-Geneidy 2017; Pereira et al. 2021),
green spaces (Reyes, Paez, and Morency 2014; Rojas et al. 2016; Liang,
Yan, and Yan 2024), schools (e.g., Williams and Wang 2014; Romanillos
and Garcia-Palomares 2018; Marques, Wolf, and Feitosa 2021), social
contacts (e.g., Neutens et al. 2007; S. Farber, Páez, and Morency 2012;
S. Farber et al. 2013), and regional economic analysis (e.g., R.
Vickerman, Spiekermann, and Wegener 1999; Lopez, Gutierrez, and Gomez
2008; Ribeiro, Antunes, and Páez 2010; Gutierrez et al. 2011) among many
other domains of application. In other words, accessibility analysis is
used today to broadly understand the potential to reach (or spatially
interact with) opportunities that are important to people (Ferreira and
Papa 2020). However, despite its growth in popularity in scholarly
works, challenges remain with respect to the more widespread adoption of
accessibility in planning practice. For instance, the diversity of
accessibility definitions has been flagged by van Wee (2016), S. Handy
(2020), and Kapatsila et al. (2023). Further, difficulties in the
interpretability and communicability of outputs has also been noticed by
many authores, including Karst T. Geurs and van Wee (2004), van Wee
(2016), and Ferreira and Papa (2020).

The adoption of accessibility in planning practice is not necessarily
made easier when potential adopters have to contend with a plethora of
definitions, each seemingly more sophisticated but less intuitive than
the last (Kapatsila et al. 2023). At a high level, Karst T. Geurs and
van Wee (2004) identify four families of accessibility measures:
infrastructure-, place-, person-, and utility-based. Of the place-based
family, which is the focus of this paper, the menu has grown to include
gravity-based accessibility (e.g., Hansen 1959; Pirie 1979), cumulative
opportunities (e.g., Wachs and Kumagai 1973; Pirie 1979; Ye et al.
2018), modified gravity (e.g., Schuurman, Berube, and Crooks 2010),
2-Step Floating Catchment Areas (e.g., Luo and Wang 2003), Enhanced
2-Step Floating Catchment Areas (e.g., Luo and Qi 2009), 3-Stage
Floating Catchment Areas (e.g., Wan, Zou, and Sternberg 2012), Modified
2-Step Floating Catchment Areas (e.g., Delamater 2013), inverse 2-Step
Floating Catchment Areas (e.g., F. Wang 2021), and n-steps Floating
Catchment Areas (Liang, Yan, and Yan 2024). How is a practitioner to
choose among this myriad options? What differences in accessibility
scores should matter, and how should they be communicated? {[}see van
Wee (2016); p.~14{]}.

Here, we seek to address the breath of place-based accessibility's
definitions and lack of interpretability by demonstrating that
mobility-oriented models--such as the commonly used gravity-based
accessibility (e.g., Hansen 1959) and the spatial interaction model
(e.g., Wilson 1971)-- are rooted in the same spatial interaction
modeling foundation. In fact, we propose that accessibility can be
specified using spatial interaction principles as a \emph{family of
measures}, akin to the family of spatial interaction models introduced
in Wilson (1971) that can be defined using balancing factors that
constrains values based on known information about the land-use.

This work aims to offer three contributions: (1) it introduces a family
of accessibility measures within the principles of spatial interaction;
(2) it formally defines three \emph{constrained} accessibility measures
by reintroducing Wilson-analogous balancing factors. These measures are
the total constrained accessibility measure, the singly constrained
accessibility measure, and the doubly constrained measure, which are
explicitly connected to popular measures such as the Hansen-type
accessibility (Hansen 1959), the popular competition approach of the
2-Step Floating Catchment Area (2SFCA) (Q. Shen 1998a; Luo and Wang
2003), and the concept of market potential (C. D. Harris 1954; R. W.
Vickerman 1974). These Wilson-analogous balancing factors introduced are
also defined in order of increasing restrictiveness to shed light on the
role of \emph{potential} in accessibility and access. (3) This work also
demonstrates that the introduction of balancing factors makes
accessibility measures easier to interpret and to communicate by
restoring the measurement units to the resulting raw accessibility
values. Each zonal and zonal flow value from a constrained accessibility
measure is always in interpretable units, namely the number of
`opportunities for spatial interaction' or `population for spatial
interaction'. This is in contrast to conventional accessibility
measures, particularly gravity based measures, that yield values in
units of `opportunities weighted by some representation of travel
friction'.

To achieve these stated objectives, this paper contends that
accessibility research must reconnect with its spatial interaction
origins. Particularly, we argue that an important aspect of spatial
interaction modelling--namely, constraining the results to match
empirical observations--was never effectively reincorporated into
accessibility analysis. Empirical constraints were embraced by early
spatial interaction literature following the work of Wilson (1971), but
this stream of literature tended to flow separately from research
inspired by Hansen (1959)' accessibility. The application of Wilson
(1971)`s empirical constraints supported the development of various
spatial interaction models that remain relevant in research and practice
today (Ortúzar and Willumsen 2011). However, the same cannot be said of
the contemporary accessibility literature, where empirical constraints
were not explicitly adopted. We argue that the absence of empirical
constraints (and their attendant proportionality constants) has
contributed to some of accessibility analysis' interpretability issues;
for instance, the fuzziness of insights beyond simple proportional
statements like `higher-than' or `lower-than' (E. J. Miller 2018).
Moreover, without constraints we lose track of what accessibility
measures, that is, the number of opportunities that can be spatially
interacted with. Without a clear sense of measurement units,
comparability between accessibility measures, across cities, and
transport modes, may also become compromised.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%

\PassOptionsToPackage{table}{xcolor}

\documentclass[
  10pt,
  letterpaper,
]{article}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% marvosym package for additional characters
\usepackage{marvosym}

% cite package, to clean up citations in the main text. Do not remove.
% Using natbib instead
% \usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% create "+" rule type for thick vertical lines
\newcolumntype{+}{!{\vrule width 2pt}}

% create \thickcline for thick horizontal lines of variable length
\newlength\savedwidth
\newcommand\thickcline[1]{%
  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
  \cline{#1}%
  \noalign{\vskip\arrayrulewidth}%
  \noalign{\global\arrayrulewidth\savedwidth}%
}

% \thickhline command for thick horizontal lines that span the table
\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
\hline
\noalign{\global\arrayrulewidth\savedwidth}}

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in 
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Header and Footer with logo
\usepackage{lastpage,fancyhdr}
\usepackage{epstopdf}
%\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
%\setlength{\headheight}{27.023pt}
%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\today}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\usepackage{colortbl}
\usepackage{array}
\usepackage{anyfontsize}
\usepackage{multirow}
\usepackage{float}
\floatplacement{table}{H}
\floatplacement{figure}{H}
\usepackage{pdflscape}
\newcommand{\blandscape}{\begin{landscape}}
\newcommand{\elandscape}{\end{landscape}}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\usepackage[numbers,square,comma]{natbib}
\bibliographystyle{plos2015}
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={A family of accessibility measures derived from spatial interaction principles},
  pdfauthor={Anastasia Soukhov; Rafael H. M. Pereira; Christopher D. Higgins; Antonio Páez},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\begin{document}
\vspace*{0.2in}

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{A family of accessibility measures derived from spatial
interaction
principles} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
\\
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
Anastasia Soukhov\textsuperscript{1}, Rafael H. M.
Pereira\textsuperscript{2}, Christopher D.
Higgins\textsuperscript{3*}, Antonio Páez\textsuperscript{1}
\\
\bigskip
\textbf{1} McMaster University, School of Earth, Environment \&
Society, Hamilton, Canada, \\ \textbf{2} Institute for Applied Economic
Research - Ipea, Data Science Division, Brasília,
Brazil, \\ \textbf{3} University of Toronto Scarborough, Department of
Human Geography, Toronto, Canada, 
\bigskip

% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
% 
% Remove or comment out the author notes below if they aren't used.
%
% Primary Equal Contribution Note
\Yinyang These authors contributed equally to this work.

% Additional Equal Contribution Note
% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
%\ddag These authors also contributed equally to this work.

% Current address notes
\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
% \textcurrency b Insert second current address 
% \textcurrency c Insert third current address

% Deceased author note
\dag Deceased

% Group/Consortium Author Note
\textpilcrow Membership list can be found in the Acknowledgments
sections

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* cd.higgins@utoronto.ca

\end{flushleft}

\section*{Abstract}
Transportation planning has long prioritized the efficiency of movement,
or mobility. However, the concept of accessibility represents a more
comprehensive evolution, shifting focus from mere movement to the
potential to reach (i.e., spatially interact) with desired destinations.
Despite growing recognition of accessibility-based planning approaches,
the concept remains fragmented, with inconsistent definitions and
unclear interpretations. This work's aim is to clarify and unify the
concept of accessibility by connecting it into spatial interaction
modeling. We demonstrate that widely used mobility and accessibility
models, such as gravity-based accessibility and spatial interaction
models, share common theoretical roots. From this foundation, this paper
offers three contributions: (A) we introduce a family of accessibility
measures within the principles of spatial interaction, and (B) formally
define four members of the family, namely the `unconstrained' measure
(i.e., Hansen-type accessibility), the `total constrained' measure
(i.e., a constrained version of the Hansen-type accessibility), the
`singly constrained' measure (i.e., related to the popular two step
floating catchment approach - 2SFCA), and the `doubly constrained'
measure representing realized interactions or `access', effectively
equal to the doubly constrained spatial interaction model; and (C) we
demonstrate the interpretability advantages of the family, as these
constrained accessibility measures yield values in units of the number
of potential ``opportunities for spatial interaction'' or ``population
for spatial interaction'' for each zone and zonal flow. The family of
accessibility measures proposed here clarifies the concept of
`potential' in accessibility, demonstrates theoretical and formulaic
linkages across popular accessibility and spatial interaction models,
and reintroduces measurement units into accessibility measures. By doing
so, we believe this family of measures can unlock a clearer, more
interpretable, and cohesive foundation for accessibility analysis.

\linenumbers

@@

In the early nineteenth century, industrializing cities grew rapidly: so
did traffic congestion and the development of transportation planning
practice focused primarily on mobility. In this newly founded practice,
access to destinations was treated as a by-product of movement.
Following the rise of the automobile and significant investments in
transportation infrastructure after World War II, this mobility-oriented
approach has led to some problematic outcomes. Specifically, the car
became seen as the ultimate mobility tool, helping to foster the
development of low-density, single-use residential neighborhoods and
entrenching an automobility mono-culture within transportation systems
\citep{miller_collaborative_2011, lavery_driving_2013}. Decades of
planning for this automobility mono-culture, still often characterized
by road and highway expansion, have been marked by increased travel
costs and environmental burdens, with limited impact on enhancing the
ease with which people can reach destinations
\citep{farber_running_2011, handyACCESSIBILITYVSMOBILITYENHANCING2002, paez_healthcare_2010}.
In response, transportation researchers have increasingly advocated for
the adoption of accessibility as a planning criterion, in contrast to
traditional mobility-oriented transportation planning approaches which
translate into indicators that benchmark movement (e.g., vehicle
kilometres traveled, intersection through traffic, etc.) which are not
necessarily linked to improved accessibility
\citep{silvaAccessibilityInstrumentsPlanning2017, paez_developing_2013, handy2020, elgeneidyMakingAccessibilityWork2022}.

Accessibility may be defined as the ``potential of opportunities for
{[}spatial{]} interaction'' \citep{hansen1959}, in contrast to mobility
which is simply, movement in itself or `spatial interaction'. Mobility
has been the basis of traditional transportation planning approaches
\citep{ortuzar_2011_modelling}. While accessibility brings a more
holistic understanding of combined transportation and land use systems,
incorporating both the explicit consideration of destination as
opportunity and the concept of \emph{potential}
\citep{handyMeasuringAccessibilityExploration1997}.

The growing interest in accessibility has been accompanied by a boom in
scholarly research using different methods and focusing on different
research contexts; it has grown to include studies of access to
employment
\citep[e.g.,][]{karstEvaluationAccessibilityImpacts2003, grengs2010job, paez_jobs_2013, merlin2017competition, tao_investigating_2020},
health care
\citep[e.g.,][]{luo2003, paez_healthcare_2010, wan2012three, delamater2013spatial, boisjoly2017informality, pereira_2021_geographic},
green spaces
\citep{reyesAccessibility2014, rojas_accessibility_2016, liang_novel_2024},
schools
\citep[e.g.,][]{williams_disparities_2014, romanillosAccessibilitySchoolsSpatial2018, marques_accessibility_2021},
social contacts
\citep[e.g.,][]{neutens_human_2007, farberActivitySpacesMeasurement2012, farber_2013_social},
and regional economic analysis
\citep[e.g.,][]{vickermanAccessibility1999, lopezMeasuring2008, ribeiro_road_2010, gutierrez_evaluating_2011}
among many other domains of application. In other words, accessibility
analysis is used today to broadly understand the potential to reach (or
spatially interact with) opportunities that are important to people
\citep{ferreiraReenactingMobilityAccessibility2020}. However, despite
its growth in popularity in scholarly works, challenges remain with
respect to the more widespread adoption of accessibility in planning
practice. For instance, the diversity of accessibility definitions has
been flagged by van Wee \citep{vanweeAccessible2016}, Handy
\citep{handy2020}, and Kapatsila et al.
\citep{kapatsila_resolving_2023}. Further, difficulties in the
interpretability and communicability of outputs has also been noticed by
many authores, including Geurs and van Wee
\citep{geursAccessibilityEvaluationLanduse2004}, van Wee
\citep{vanweeAccessible2016}, and Ferreira and Papa
\citep{ferreiraReenactingMobilityAccessibility2020}.

The adoption of accessibility in planning practice is not necessarily
made easier when potential adopters have to contend with a plethora of
definitions, each seemingly more sophisticated but less intuitive than
the last \citep{kapatsila_resolving_2023}. At a high level, Geurs and
van Wee \citep{geursAccessibilityEvaluationLanduse2004} identify four
families of accessibility measures: infrastructure-, place-, person-,
and utility-based. Of the place-based family, which is the focus of this
paper, the menu has grown to include gravity-based accessibility
\citep[e.g.,][]{hansen1959, pirie_measuring_1979}, cumulative
opportunities
\citep[e.g.,][]{wachs_physical_1973, pirie_measuring_1979, ye_spatial_2018},
modified gravity \citep[e.g.,][]{SchuurmanMeasuring2010}, 2-Step
Floating Catchment Areas \citep[e.g.,][]{luo2003}, Enhanced 2-Step
Floating Catchment Areas \citep[e.g.,][]{luoEnhanced2009}, 3-Stage
Floating Catchment Areas \citep[e.g.,][]{wan2012three}, Modified 2-Step
Floating Catchment Areas \citep[e.g.,][]{delamater2013spatial}, inverse
2-Step Floating Catchment Areas \citep[e.g.,][]{wang_2sfca_2021}, and
n-steps Floating Catchment Areas \citep{liang_novel_2024}. How is a
practitioner to choose among this myriad options? What differences in
accessibility scores should matter, and how should they be communicated?
\citep[see][ p.~14]{vanweeAccessible2016}.

Here, we seek to address the breath of place-based accessibility's
definitions and lack of interpretability by demonstrating that
mobility-oriented models--such as the commonly used gravity-based
accessibility \citep[e.g.,][]{hansen1959} and the spatial interaction
model \citep[e.g.,][]{wilson1971}-- are rooted in the same spatial
interaction modeling foundation. In fact, we propose that accessibility
can be specified using spatial interaction principles as a \emph{family
of measures}, akin to the family of spatial interaction models
introduced in Wilson \citep{wilson1971} that can be defined using
balancing factors that constrains values based on known information
about the land-use.

This work aims to offer three contributions: (1) it introduces a family
of accessibility measures within the principles of spatial interaction;
(2) it formally defines three \emph{constrained} accessibility measures
by reintroducing Wilson-analogous balancing factors. These measures are
the total constrained accessibility measure, the singly constrained
accessibility measure, and the doubly constrained measure, which are
explicitly connected to popular measures such as the Hansen-type
accessibility \citep{hansen1959}, the popular competition approach of
the 2-Step Floating Catchment Area (2SFCA)
\citep{shenLocationCharacteristicsInnercity1998, luo2003}, and the
concept of market potential
\citep{harris_market_1954, vickermanAccessibilityAttractionPotential1974}.
These Wilson-analogous balancing factors introduced are also defined in
order of increasing restrictiveness to shed light on the role of
\emph{potential} in accessibility and access. (3) This work also
demonstrates that the introduction of balancing factors makes
accessibility measures easier to interpret and to communicate by
restoring the measurement units to the resulting raw accessibility
values. Each zonal and zonal flow value from a constrained accessibility
measure is always in interpretable units, namely the number of
`opportunities for spatial interaction' or `population for spatial
interaction'. This is in contrast to conventional accessibility
measures, particularly gravity based measures, that yield values in
units of `opportunities weighted by some representation of travel
friction'.
